---
title: "Practicum II - Part III: Analyze Data"
author: "Connor Clancy - clancy.co@northeastern.edu"
date: "Spring 2023"
output: pdf_document
---

## Libary Imports
```{r libraryImports}
library(RMySQL)
library(sqldf)
options(sqldf.driver = "SQLite")
```

```{r mysqlConnection}
# MySQL Connection
db_user <- 'admin'
db_password <- 'Northea$tern23'
db_name <- 'practicumTwo'
db_host <- 'cclancy-cs5200.cbowkysg1oyc.us-east-2.rds.amazonaws.com'
db_port <- 3306

dbcon <- dbConnect(MySQL(), user = db_user, password = db_password,
                   dbname = db_name, host = db_host, port = db_port)
```

## Analytical Query I 
_*Top five journals with the most articles published in them for the time period. Think about how you would best present this information so it is easily usable by an analyst or manager.*_

```{sql a1, connection=dbcon}
SELECT 
  jid, 
	journal_title, 
	SUM(articles) AS articles_published
FROM factJournals
GROUP BY 
  jid, 
  journal_title 
ORDER BY 
  SUM(articles) DESC
LIMIT 5
```

The chart above shows the Journals from the data set that published the most articles between January 1975 and  January 1979. 

## Analytical Query II 
_*Number of articles per journal per year broken down by quarter. Again, think of a good way to show this. Is a table better than a visualization or is it better to use a visualization.*_

```{sql a2SQL, connection=dbcon}
SELECT 
	jid, 
  journal_title, 
  SUM(CASE WHEN pub_year = 1975 AND pub_quarter = 1 THEN articles ELSE 0 END) AS '1975-1',
  SUM(CASE WHEN pub_year = 1975 AND pub_quarter = 2 THEN articles ELSE 0 END) AS '1975-2',
  SUM(CASE WHEN pub_year = 1975 AND pub_quarter = 3 THEN articles ELSE 0 END) AS '1975-3',
  SUM(CASE WHEN pub_year = 1975 AND pub_quarter = 4 THEN articles ELSE 0 END) AS '1975-4',
	SUM(CASE WHEN pub_year = 1976 AND pub_quarter = 1 THEN articles ELSE 0 END) AS '1976-1',
  SUM(CASE WHEN pub_year = 1976 AND pub_quarter = 2 THEN articles ELSE 0 END) AS '1976-2',
  SUM(CASE WHEN pub_year = 1976 AND pub_quarter = 3 THEN articles ELSE 0 END) AS '1976-3',
  SUM(CASE WHEN pub_year = 1976 AND pub_quarter = 4 THEN articles ELSE 0 END) AS '1976-4',
  SUM(CASE WHEN pub_year = 1977 AND pub_quarter = 1 THEN articles ELSE 0 END) AS '1977-1',
  SUM(CASE WHEN pub_year = 1977 AND pub_quarter = 2 THEN articles ELSE 0 END) AS '1977-2',
  SUM(CASE WHEN pub_year = 1977 AND pub_quarter = 3 THEN articles ELSE 0 END) AS '1977-3',
  SUM(CASE WHEN pub_year = 1977 AND pub_quarter = 4 THEN articles ELSE 0 END) AS '1977-4',
	SUM(CASE WHEN pub_year = 1978 AND pub_quarter = 1 THEN articles ELSE 0 END) AS '1978-1',
  SUM(CASE WHEN pub_year = 1978 AND pub_quarter = 2 THEN articles ELSE 0 END) AS '1978-2',
  SUM(CASE WHEN pub_year = 1978 AND pub_quarter = 3 THEN articles ELSE 0 END) AS '1978-3',
  SUM(CASE WHEN pub_year = 1978 AND pub_quarter = 4 THEN articles ELSE 0 END) AS '1978-4',
  SUM(CASE WHEN pub_year = 1979 AND pub_quarter = 1 THEN articles ELSE 0 END) AS '1979-1',
  SUM(articles) AS total_articles
FROM factJournals 
WHERE pub_year IS NOT NULL AND pub_quarter IS NOT NULL
GROUP BY 
	jid, 
    journal_title
ORDER BY 
	SUM(articles) DESC
```

Th above chart shows all of the articles published by each journal for each quarter of the data set.  As it may be more useful to view the data trends over time, an additional chart has been provided below which shows the aggregate trends in articles being published by quarter below. 

```{r a2}
df_chart_raw <- dbGetQuery(
  dbcon, 
  "
  SELECT 
  	pub_year, 
  	SUM(CASE WHEN pub_quarter = 1 THEN articles ELSE 0 END) AS q1,
    SUM(CASE WHEN pub_quarter = 2 THEN articles ELSE 0 END) AS q2,
    SUM(CASE WHEN pub_quarter = 3 THEN articles ELSE 0 END) AS q3,
    SUM(CASE WHEN pub_quarter = 4 THEN articles ELSE 0 END) AS q4
  FROM factJournals 
  WHERE pub_year IS NOT NULL AND pub_quarter IS NOT NULL
  GROUP BY pub_year
  ORDER BY pub_year;
  "
)

df_chart <- df_chart_raw[,-1]
rownames(df_chart) <- df_chart_raw[,1]

colors <- c("red", "blue", "yellow", "darkgreen")

barplot(
  t(as.matrix(df_chart)), beside=TRUE, 
  xlab = "Year", 
  ylab = " Articles", 
  col = colors, 
  main = "Published Articles"
  )
legend("topright", c("Q1", "Q2", "Q3", "Q4"), fill = colors)
```
```{r disconnect}
dbDisconnect(dbcon)
```









