---
title: 'Practicum I - Bird Strike Database'
author: 
- 'Connor Clancy'
date: 'Spring 2023'
---

## Connect to Database
```{r database_connection}
library(RMySQL)
library(sqldf)
options(sqldf.driver = "SQLite")

db_user <- 'admin'
db_password <- 'Northea$tern23'
db_name <- 'practicumOne'
db_host <- 'cclancy-cs5200.cbowkysg1oyc.us-east-2.rds.amazonaws.com'
db_port <- 3306

dbcon <- dbConnect(MySQL(), user = db_user, password = db_password,
                   dbname = db_name, host = db_host, port = db_port)
```

## Create Helper R Functions 
```{r helperFunctions}
#' Method to check the existence of a foreign key and delete it if it exists.
#' If a foreign key already exists in a database this can cause problems for
#' dropping or altering the database.
#' 
#' @param fk_name the name of the foreign key to be checked
#' @return void
#' 
delete_fk <- function(fk_name) {
  # Query the database for the foreign key
  iac <- dbGetQuery(dbcon, 
    sprintf(
      "SELECT TRUE
      FROM information_schema.table_constraints
      WHERE constraint_name = '%s';", fk_name
    )
  )
  
  # If the foreign key exist, delete it
  if (nrow(iac) > 0) {
    dbExecute(dbcon, 
      sprintf(
        "ALTER TABLE incidents DROP FOREIGN KEY %s;", fk_name
      )
    )
  }  
}

#' Method to take in a date string formatted m/d/yyyy and return it as a 
#' properly formatted R date. 
#' 
#' @param ds the date string to be transformed
#' @return date formatted version of the input string.
#' 
format_date <- function(ds) {
  td <- strsplit(ds, "/")[[1]]
  return(paste(td[3],td[1],td[2], sep="/"))
}
```

## Create Database (Task 4)

### 4A - Incidents
```{sql dropTableIncidents, connection=dbcon}
DROP TABLE IF EXISTS incidents;
```

```{sql createTableIncidents, connection=dbcon}
CREATE TABLE incidents (
  rid INT AUTO_INCREMENT PRIMARY KEY,
  `dep.date` DATE,
  origin INT,
  airline INT, 
  aircraft VARCHAR(255) NOT NULL,
  `flight.phase` ENUM('takeoff', 'landing', 'inflight', 'unknown'),
  altitude INT,
  conditions INT, 
  warned TINYINT,
  CHECK(altitude>=0)
);
```

### 4B - Airports
```{sql dropTableAirports, connection=dbcon}
DROP TABLE IF EXISTS airports;
```

```{sql createTableAirports, connection=dbcon}
CREATE TABLE airports(
  aid INT AUTO_INCREMENT PRIMARY KEY,
  airportName VARCHAR(255) NOT NULL,
  airportCode VARCHAR(3),
  state VARCHAR(255)
);
```

### 4C - Foreign Key Incident Airports 
```{r dropFKIncidentsAirports}
delete_fk('FK_IncidentAirports')
```

```{sql connection=dbcon}
ALTER TABLE incidents
ADD CONSTRAINT FK_IncidentAirports 
FOREIGN KEY (origin) REFERENCES airports(aid);
```

### 4D - Conditions 
```{sql dropTableConditions, connection=dbcon}
DROP TABLE IF EXISTS conditions;
```

```{sql createTableConditions, connection=dbcon}
CREATE TABLE conditions(
  cid INT AUTO_INCREMENT PRIMARY KEY,
  `condition` VARCHAR(255) NOT NULL,
  explanation VARCHAR(255)
);
```

```{r dropFKIncidentsConditions}
delete_fk('FK_IncidentConditions')
```

```{sql dropIncidentConditionFK, connection=dbcon}
ALTER TABLE incidents
ADD CONSTRAINT FK_IncidentConditions 
FOREIGN KEY (conditions) REFERENCES conditions(cid);
```

### 4E - Airlines
```{sql dropTableAirlines, connection=dbcon}
DROP TABLE IF EXISTS airlines;
```

```{sql createTableAirlines, connection=dbcon}
CREATE TABLE airlines(
  eid INT AUTO_INCREMENT PRIMARY KEY,
  airlineName VARCHAR(255) NOT NULL,
  airlineCode VARCHAR(2),
  flag VARCHAR(255)
);
```

### 4F - Join Incidents Airlines 
```{r dropFKIncidentsAirlines}
delete_fk('FK_IncidentAirline')
```

```{sql dropFKIncidentsAirlinesFK, connection=dbcon}
ALTER TABLE incidents
ADD CONSTRAINT FK_IncidentAirline 
FOREIGN KEY (airline) REFERENCES airlines(eid);
```

### 4G - Test Code to ensure the tables are properly set up 

#### Test Inserts 

##### Airline Table Tests
```{sql testAirlinesInsert, connection=dbcon, eval=F}
INSERT INTO airlines (airlineName, airlineCode) VALUES 
  ('American Airlines', 'AA'),
  ('Delta', 'DL'),
  ('jetBlue', 'B6'),
  ('Southwest', 'WN'),
  ('United', 'UA');
```

```{sql testAirlinesSelect, connection=dbcon, eval=F}
SELECT 
  * 
FROM airlines;
```

##### Conditions Table Tests
```{sql testConditionsInsert, connection=dbcon, eval=F}
INSERT INTO conditions (`condition`, explanation) VALUES 
  ('clear', 'clear skies'),
  ('overcast', 'light clounds');
```

```{sql testConditionsSelect, connection=dbcon, eval=F}
SELECT 
  * 
FROM conditions;
```

##### Airport Table Tests
```{sql testAirportInserts, connection=dbcon, eval=F}
INSERT INTO airports (airportName, airportCode, state) VALUES 
  ('Boston Logan International Airport', 'BOS', 'MA'), 
  ('Seattle-Tacoma International Airport', 'SEA', 'WA'),
  ('Vancouver International Airport', 'YVR', 'BC');
```

```{sql testAirportSelect, connection=dbcon, eval=F}
SELECT 
  * 
FROM airports;
```

##### Incidents Table Tests
```{sql testIncidentsInsert, connection=dbcon, eval=F}
INSERT INTO incidents (`dep.date`, origin, airline, aircraft, `flight.phase`, altitude, conditions, warned) VALUES 
  ('2023-02-24', 1, 3, 'A321-200', 'takeoff', 32, 2, 1),
  ('2023-02-26', 2, 1, '737 MAX 9', 'landing', 189, 1, 0);
```

```{sql testIncidentSelect, connection=dbcon, eval=F}
SELECT 
  * 
FROM incidents;
```

##### Clean-Up Test Data 
```{sql testIncidentsDelete, connection=dbcon, eval=F}
DELETE FROM incidents WHERE 1=1;
```

```{sql testAirlineDelete, connection=dbcon, eval=F}
DELETE FROM airlines WHERE 1=1;
```

```{sql testConditionDelete, connection=dbcon, eval=F}
DELETE FROM conditions WHERE 1=1;
```

```{sql testAirportsDelete, connection=dbcon, eval=F}
DELETE FROM airports WHERE 1=1;
```

## Load Data (Task 5)
```{r readRawCSV}
bds.raw <- read.csv("BirdStrikesData-V2.csv", header = TRUE)
head(bds.raw)
```

## Populate Tables (Task 6)

### Airlines Data Population
```{r prepareAirlineDF}
# TRIM() function included to remove a typo in the data.
bds.airlines <- sqldf("
                      SELECT DISTINCT 
                        0 AS eid,
                        trim(airline, '*') AS airlineName,
                        NULL AS airlineCode, 
                        NULL AS flag
                      FROM `bds.raw` 
                      WHERE TRIM(airline, ' ') != ''
                      ORDER BY trim(airline, '*')
                      ")

# Create the artificial primary key with a counter 
n.airlines <- nrow(bds.airlines)
bds.airlines[,1] <- seq(1, n.airlines)

# Display the data frame to make sure it looks correct
head(bds.airlines)
```

```{r writeAirlineTable}
dbWriteTable(dbcon, "airlines", bds.airlines, overwrite = F, append = T, row.names = FALSE)
```

### Airports Data Population
```{r prepareAirportDF}
# Retrieve the relevant data from the raw dataframe.
bds.airports <- sqldf("
                      SELECT DISTINCT 
                        0 AS aid,
                        airport AS airportName,
                        NULL AS airportCode, 
                        CASE 
                          WHEN origin = 'N/A' THEN NULL 
                          ELSE origin 
                        END AS state
                      FROM `bds.raw` 
                      WHERE TRIM(airport, ' ') != ''
                      ORDER BY airport
                      ")

# Create the artificial primary key with a counter 
n.airports <- nrow(bds.airports)
bds.airports[,1] <- seq(1, n.airports)

# Display the data frame to make sure it looks correct
head(bds.airports)
```

```{r writeAirportTable}
dbWriteTable(dbcon, "airports", bds.airports, overwrite = F, append = T, row.names = FALSE)
```

### Condition Data Population
```{r prepareConditionsDF}
# Retrieve the relevant data from the raw data frame.
bds.conditions <- sqldf("
                      SELECT DISTINCT 
                        0 AS cid,
                        sky_conditions AS condition,
                        NULL AS explanation 
                      FROM `bds.raw` 
                      ORDER BY sky_conditions
                      ")

# Create the artificial primary key with a counter 
n.conditions <- nrow(bds.conditions)
bds.conditions[,1] <- seq(1, n.conditions)

# Display the data frame to make sure it looks correct
head(bds.conditions)
```

```{r writeConditionsTable}
dbWriteTable(dbcon, "conditions", bds.conditions, overwrite = F, append = T, row.names = FALSE)
```

### Incidents Data Population
```{r prepareIncidentsDF}
# Retrieve the relevant data from the raw data frame.
bds.incidents <- sqldf("
                      SELECT 
                        0 AS rid,
                        REPLACE(flight_date, ' 0:00', '') AS `dep.date`, 
                        airport AS origin, 
                        trim(airline, '*') AS airline, 
                        model AS aircraft, 
                        flight_phase AS `flight.phase`, 
                        CAST(REPLACE(altitude_ft, ',','') AS INT) AS altitude,
                        sky_conditions AS conditions, 
                        CASE
                          WHEN pilot_warned_flag = 'Y' THEN 1 
                          ELSE 0 
                        END AS warned
                      FROM `bds.raw` 
                      ")

# Create the artificial primary key with a counter 
n.incidents <- nrow(bds.incidents)
bds.incidents[,1] <- seq(1, n.incidents)

# Format date column
for (r in 1:n.incidents) {
  bds.incidents$dep.date[r] = format_date(bds.incidents$dep.date[r])
}

# Link `airline` foreign key 
for (r in 1:n.incidents) {
  eid <- bds.airlines$eid[which(bds.airlines$airlineName == bds.incidents$airline[r])]
  
  # incidents with a blank airline get a NULL value
  if(length(eid) != 0) {
    bds.incidents$airline[r] <- eid
  } else {
    bds.incidents$airline[r] <- NA
  }
}

# Link `airport` foreign key 
for (r in 1:n.incidents) {
  aid <- bds.airports$aid[which(bds.airports$airportName == bds.incidents$origin[r])]
  
  # incidents with a blank airports get a NULL value
  if(length(aid) != 0) {
    bds.incidents$origin[r] <- aid
  } else {
    bds.incidents$origin[r] <- NA
  }
}

# Link `condition` foreign key 
for (r in 1:n.incidents) {
  cid <- bds.conditions$cid[which(bds.conditions$condition == bds.incidents$conditions[r])]
  
  # incidents with a blank condition get a NULL value
  if(length(cid) != 0) {
    bds.incidents$conditions[r] <- cid
  } else {
    bds.incidents$conditions[r] <- NA
  }
}

# Display the data frame to make sure it looks correct
head(bds.incidents)
```

```{r writeIncidentsTable}
dbWriteTable(dbcon, "incidents", bds.incidents, overwrite = F, append = T, row.names = FALSE)
```


## Show Table Heads (Task 7)

## States with the most bird strikes (Task 8)

## Airlines with above average bird strikes (Task 9)

## Birdstrikes by month and fligth phase (Task 10)

## Birdstrike Scatter Plot (Task 11)

## Stored Procedure for adding an incident to the database (Task 12)

